\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-22 paper.qmd
Update true model screenshots and youtube link

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional manifold, tour, data vizualization, model in the data space]
keywords-formatted: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional manifold, tour, data vizualization, model in the data space]

@@

Three views of the true model in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/tYATMiSlCXQ>). 
:::

XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others.

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Austria
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional (high-D) data using non-linear transformation. The methods and parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of high-D distributions. To help evaluate the NLDR we have developed an algorithm to show the 2D NLDR model in the high-D space, viewed with a tour. One can see if the model fits everywhere or better in some subspaces, or completely mismatches the data. It is used to evaluate which 2D layout is the best representation of the high-D distribution and see how different methods may have similar summaries or quirks.

keywords: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data space]
keywords-formatted: [high-dimensional data, dimension reduction, hexagon binning, low-dimensional representation, tour, data vizualization, model-in-the-data space]

@@

Three views of the true model in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape. (The **langevitour** software is used to view the data with a tour, and the full video is available at <https://youtu.be/I5GL23vLiw0>). 
:::

<!-- XXX The points on the "theoretical model" should be equidistant along the S. They are still to bunched in some spots and too far apart in others. -->

### 2024-07-22 paper.tex
Update true model screenshots and youtube link

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension reduction, hexagon
binning, low-dimensional manifold, tour, data vizualization, model in
the data space},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\bigskip
\bigskip
\begin{abstract}
Nonlinear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional (high-D)
data using non-linear transformation. The methods and parameter choices
can create wildly different representations, making it difficult to
decide which is best, or whether any or all are accurate or misleading.
NLDR often exaggerates random patterns, sometimes due to the samples
observed. But NLDR views have an important role in data analysis
because, if done well, they provide a concise visual (and conceptual)
summary of high-D distributions. To help evaluate the NLDR we have
developed an algorithm to show the 2D NLDR model in the high-D space,
viewed with a tour. One can see if the model fits everywhere or better
in some subspaces, or completely mismatches the data. It is used to
evaluate which 2D layout is the best representation of the high-D
distribution and see how different methods may have similar summaries or
quirks.
\end{abstract}

\noindent%
{\it Keywords:} high-dimensional data, dimension reduction, hexagon
binning, low-dimensional manifold, tour, data vizualization, model in
the data space
\vfill

@@

\caption{\label{fig-scurve-true-sc}Three views of the true model in
projections from \(7\text{-}D\), for the S-curve data. The model closely
fits the shape. (The \textbf{langevitour} software is used to view the
data with a tour, and the full video is available at
\url{https://youtu.be/tYATMiSlCXQ}).}

@@

XXX The points on the ``theoretical model'' should be equidistant along
the S. They are still to bunched in some spots and too far apart in
others.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Looking at Non-Linear Dimension Reductions as Models in the Data Space},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension reduction, hexagon
binning, low-dimensional representation, tour, data
vizualization, model-in-the-data space},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\bigskip
\bigskip
\begin{abstract}
Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional (high-D)
data using non-linear transformation. The methods and parameter choices
can create wildly different representations, making it difficult to
decide which is best, or whether any or all are accurate or misleading.
NLDR often exaggerates random patterns, sometimes due to the samples
observed. But NLDR views have an important role in data analysis
because, if done well, they provide a concise visual (and conceptual)
summary of high-D distributions. To help evaluate the NLDR we have
developed an algorithm to show the 2D NLDR model in the high-D space,
viewed with a tour. One can see if the model fits everywhere or better
in some subspaces, or completely mismatches the data. It is used to
evaluate which 2D layout is the best representation of the high-D
distribution and see how different methods may have similar summaries or
quirks.
\end{abstract}

\noindent%
{\it Keywords:} high-dimensional data, dimension reduction, hexagon
binning, low-dimensional representation, tour, data
vizualization, model-in-the-data space
\vfill

@@

\caption{\label{fig-scurve-true-sc}Three views of the true model in
projections from \(7\text{-}D\), for the S-curve data. The model closely
fits the shape. (The \textbf{langevitour} software is used to view the
data with a tour, and the full video is available at
\url{https://youtu.be/I5GL23vLiw0}).}

@@
